HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-13 scnn.tex
method: graph fourier basis

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}} % the laplacian matrix
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

From the HEALPix pixelization, we build a weighted undirected graph $\G = (\V, \E, \W)$, where $\V$ is the set of $N_{pix} = |\V|$ nodes, $\E$ is the set of edges, and $\W$ is the weighted adjacency matrix. In our graph, each pixel $i$ is represented by a node (also called vertex) $v_i \in \V$. Each node $v_i$ is then connected to the $8$ (or $7$)\footnote{\label{neighbors}The $12 \cdot 4 = 48$ pixels at the corner of each rhombus of the base dodecahedron only have 7 neighboring pixels. See Figures~\ref{fig:healpix_graph_4} and~\ref{fig:pooling}.} nodes $v_j$ which represent the neighboring pixels $j$ of pixel $i$, forming edges $(v_i, v_j) \in \E$. Given those edges, we define the weighted adjacency matrix $\W$ as
\begin{equation*}
	\W_{ij} = \begin{cases}
		\exp \left( -\frac{\|\x_i-\x_j\|_2^2}{\sigma^2} \right) & \text{if pixels $i$ and $j$ are neighbors,} \\
		0 & \text{otherwise,} \\
	\end{cases}
\end{equation*}
where $\x_i$ is a vector encoding the 3-dimensional coordinates of pixel $i$, and
\begin{equation*}
	\sigma = \frac{1}{|\E|} \sum_{(v_i, v_j) \in \E} \|\x_i-\x_j\|_2
\end{equation*}
is the average Euclidean distance over all connected pixels. This weighting scheme is important as while pixel areas are equal, distances between pixels are not equal. \figref{healpix_graph_4} shows a graph constructed from the HEALPix sampling of a sphere.

@@

\subsection{Graph Fourier basis and spherical harmonics}

Following~\cite{shuman2013emerging}, the normalized graph Laplacian,
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$, is a second order differential operator
that can be used to define a Fourier basis on the graph. Here $\D$ is the diagonal
matrix where $\D_{ii} = \b{d}_i$ and $\b{d}_i = \sum_j \W_{ij}$ is the weighted degree of node $v_i$. By construction the Laplacian is symetric positive
semi-definite and hence can be decomposed as $\L=\U \bLambda \U^*$, where $\U$ is an
orthonormal matrix of eigenvectors and $\bLambda$, the diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors.
The graph Fourier transform of $f$ is simply its projection on $\U$ given by
$\hat{\b{f}}=\U^*\b{f}$. Similarly, the inverse graph Fourier transform reads $\b{f}=\U\hat{\b{f}}$.
Note that instead of frequencies, the Fourier modes are ordered in the increasing order of the Laplacian eigenvalues $\bLambda$. In a sense, the Laplacian
eigenvalues correspond to the squared frequencies.

As shown in \figref{graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the Healpix sampling. This topic is further discussed in Appendix~\ref{sec:spherical_harmonics}.

@@

A convolution can be defined on graphs by the convolution theorem which states that a convolution in the vertex (or time, or spatial) domain is a
multiplication in the spectral domain. Given a convolution kernel
$h: \R_+ \rightarrow \R$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{eqn:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U^* \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\eqnref{graph_convolution} contains three parts: a) $\U^* \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

@@

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$. They are a generalization of the squared frequencies in the graph setting. Hence the kernel $K_t$ can also be considered as a generalization of the
Gaussian kernel on the sphere. In \figref{gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(\figref{gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.
In~\figref{gaussian_filters_visualization} of Appendix~\ref{sec:filter_visualization}, we provide additional visualizations of the
convolution kernel $K_t$.

@@

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{The first 16 Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the Healpix sampling of the sphere. Eigenvectors 1--3 could be associated with spherical harmonics of degree $\ell=1$, eigenvectors 4--8 with degree $\ell=2$, and eigenvectors 9--15 with degree $\ell=3$. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics. \todo{show group of eigenvectors as indication of mode $\ell$. Either show eigenvalues on this plot or add a plot of the eigenvalues themselves.}}
\label{fig:graph_harmonics}
\end{figure*}

>>>>

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}}
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\bu}{\b{u}}
\newcommand{\f}{\b{f}}
\newcommand{\trans}{^\intercal}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

From the HEALPix pixelization, we build a weighted undirected graph $\G = (\V, \E, \W)$, where $\V$ is the set of $N_{pix} = |\V|$ nodes, $\E$ is the set of edges, and $\W$ is the weighted adjacency matrix. In our graph, each pixel $i$ is represented by a node (also called vertex) $v_i \in \V$. Each node $v_i$ is then connected to the $8$ (or $7$)\footnote{\label{neighbors}The $12 \cdot 4 = 48$ pixels at the corner of each rhombus of the base dodecahedron only have 7 neighboring pixels. See Figures~\ref{fig:healpix_graph_4} and~\ref{fig:pooling}.} nodes $v_j$ which represent the neighboring pixels $j$ of pixel $i$, forming edges $(v_i, v_j) \in \E$. Given those edges, we define the weighted adjacency matrix $\W \in \R^{N_{pix} \times N_{pix}}$ as
\begin{equation*}
	\W_{ij} = \begin{cases}
		\exp \left( -\frac{\|\x_i-\x_j\|_2^2}{\sigma^2} \right) & \text{if pixels $i$ and $j$ are neighbors,} \\
		0 & \text{otherwise,} \\
	\end{cases}
\end{equation*}
where $\x_i$ is a vector encoding the 3-dimensional coordinates of pixel $i$, and
\begin{equation*}
	\sigma = \frac{1}{|\E|} \sum_{(v_i, v_j) \in \E} \|\x_i-\x_j\|_2
\end{equation*}
is the average Euclidean distance over all connected pixels. This weighting scheme is important as while pixel areas are equal, distances between pixels are not equal. \figref{healpix_graph_4} shows a graph constructed from the HEALPix sampling of a sphere.

@@

\subsection{Graph Fourier basis}

Following~\cite{shuman2013emerging}, the normalized graph Laplacian,
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$, is a second order differential operator
that can be used to define a Fourier basis on the graph. Here $\D$ is the diagonal
matrix where $\D_{ii} = \b{d}_i$ and $\b{d}_i = \sum_j \W_{ij}$ is the weighted degree of node $v_i$. By construction, the Laplacian is symmetric positive
semi-definite and hence can be decomposed as $\L = \U \bLambda \U\trans$, where $\U = [\b u_1, \ldots, \b u_{N_{pix}}]$ is an
orthonormal matrix of eigenvectors and $\bLambda$ is a diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors, motivated by the fact that a Fourier basis should diagonalize the Laplacian operator.
The graph Fourier transform of a signal $\f \in \R^{N_{pix}}$ is simply its projection on the eigenvectors given by
$\hat{\f} = \mathcal{F}_\G \{\f\} = \U\trans \f$. It follows that the inverse graph Fourier transform reads $\mathcal{F}^{-1}_\G \{\hat{\f}\} = \U\hat{\f} = \U \U\trans \f = \f$.
Note that the Fourier modes are ordered in the increasing order of the Laplacian eigenvalues $\bLambda$, which can be interpreted as squared frequencies.
Indeed,
\begin{equation*}
	\bLambda_{ii} = \b u_i\trans \L \b u_i = \sum_{(v_j, v_k) \in \E} \frac{\W_{jk}}{\sqrt{\b d_j \b d_k}} (\U_{ji} - \U_{ki})^2
\end{equation*}
is a measure of how smooth the eigenvector $\b u_i$ is on the graph defined by the Laplacian $\L$.

It turns out that using the graph construction described above, the graph Fourier modes resemble the spherical harmonics (see \figref{graph_harmonics}). That is a strong hint that the graph is able to capture the spherical properties of the HEALPix sampling. This topic is further discussed in \ref{sec:spherical_harmonics}.

@@

A convolution can be defined on graphs by the convolution theorem which states that a convolution in the vertex (or time, or spatial) domain is a
multiplication in the spectral domain. Given a convolution kernel
$h: \R_+ \rightarrow \R$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{eqn:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U\trans \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\eqnref{graph_convolution} contains three parts: a) $\U\trans \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

@@

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$. They are a generalization of the squared frequencies in the graph setting. Hence the kernel $K_t$ can also be considered as a generalization of the
Gaussian kernel on the sphere. In \figref{gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(\figref{gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.
In~\figref{gaussian_filters_visualization} of Appendix~\ref{sec:filter_visualization}, we provide additional visualizations of the
convolution kernel $K_t$.

@@

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{The first 16 Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the HEALPix sampling of the sphere. Eigenvectors 1--3 could be associated with spherical harmonics of degree $\ell=1$ and order $m=(-1,0,1)$, eigenvectors 4--8 with degree $\ell=2$ and order $m=(-2,-1,0,1,2)$, and eigenvectors 9--15 with degree $\ell=3$ and order $m=(-3,\ldots,3)$. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics. \todo{show group of eigenvectors as indication of mode $\ell$. Either show eigenvalues on this plot or add a plot of the eigenvalues themselves.}}
\label{fig:graph_harmonics}
\end{figure*}
